Early environmental conditions do not impact behavioural flexibility in an invasive and non-invasive lizard species

### 2024-04-04 ms.qmd
Updates

Our early environment manipulation did not affect the lizards' performance in the task. We predicted that prenatal CORT treatment would decrease reversal learning abilities and that high temperatures would minimise these effects. However, our results show no significant differences between treatments in both species. In contrast to previous studies that show high GCs levels impaired performance in different types of learning tests [@lemaire_prenatal_2000; @zhu_prenatal_2004;  @farrell_developmental_2016], we did not find any effect of CORT on the speed at which lizards learn. Our results are, however, consistent with other experiments where GCs did not impact learning abilities [@szuran_water_1994; @bebus_associative_2016]. In some systems, exposure to CORT during development has hormetic effects on learning and neural structures [@du_dynamic_2009; @mcewen_brain_2012]. For example, in Sprague–Dawley rats, exposure to high doses of CORT, but not low doses, reduced density of immature neurons and cell proliferation and in the dentate gyrus [@DOSEneurobrummelte2010chronic]. The CORT treatment employed here is known to impact other aspects such as growth or mortality in these species (unpublished data), but the effect of the same dose can differ between tissues (@lemaire_prenatal_2000). More striking is the absence of an effect of incubation temperature since most of the studies in reptiles reported better performance when incubated at higher temperatures in different associative learning and motor tasks [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. Furthermore, those studies that reported cold-incubated lizards learn faster [@dayananda_incubation_2017; @abayarathna_effects_2020] used incubation temperatures far beyond the natural range in their hot treatments. We know that our prenatal treatments affect other traits in both species of lizards used in this experiment (unpublished data). However, it is possible that those treatments affect tissues differently. An alternative hypothesis is that the effects of the prenatal treatments are age-dependent [see @frisone2002social], but our subjucts were tested before maturity (see Supplementary material). Conversely, the effects of the early environment may likely vary depending on the cognitive task, as evidenced by studies showing variations in the impact of prenatal GCs and temperature on different brain regions [@coomber_independent_1997; @sakata_neural_2000; @lemaire_prenatal_2000]. These hypotheses are not mutually exclusive and could be tested in future experiments using multiple doses of prenatal CORT treatment, assessing learning using multiple cognitive tasks and at different ages, and measuring the neurophysiological effects on different brain nuclei.

@@

```{r, modelsage}
#| label: modelsage
# Fitting the models for L. delicata and L. guichenoti including age as a covariate
#
refit2 = TRUE
formula <- FC_reversal ~ age.start + trial_reversal*cort*temp + (1 + trial_reversal|lizard_id)
species <- c("delicata", "guichenoti")
#
models <- list() # Create an empty list to store the models
for(s in species){
  # Subset the data for the species
  data_species <- clean_df %>%
    filter(species == s)
  # Define model file name
  model_file <- paste0("output/models/model_age_", s, ".rds")
# Fit the model only if it has not been fit yet (if refit2 = TRUE)
  if(refit2){
    # Fit the model
    model <- brm(formula,
                data = data_species,
                family = bernoulli(link = "logit"),
                chains = 4, cores = 4, iter = 3000, warmup = 1000, control = list(adapt_delta = 0.99))
    # Write the model to a file
    saveRDS(model, file = model_file)
  } else {
      # Read the model from a file
      model <- readRDS(file = model_file)
  }
  # Save the model to the list
  models[[s]] <- model
}
model_age_deli <- models[["delicata"]]
model_age_guich <- models[["guichenoti"]]
```

Results for L_delicata:
` r summary(model_age_deli)`

Results for L_guichenoti:
` r summary(model_age_guich)`

>>>>

Our early environment manipulation did not affect the lizards' performance in the task. We predicted that prenatal CORT treatment would decrease reversal learning abilities and that high temperatures would minimise these effects. However, our results show no significant differences between treatments in both species. In contrast to previous studies that show high GCs levels impaired performance in different types of learning tests [@lemaire_prenatal_2000; @zhu_prenatal_2004;  @farrell_developmental_2016], we did not find any effect of CORT on the speed at which lizards learn. Our results are, however, consistent with other experiments where GCs did not impact learning abilities [@szuran_water_1994; @bebus_associative_2016]. In some systems, exposure to CORT during development has hormetic effects on learning and neural structures [@du_dynamic_2009; @mcewen_brain_2012]. For example, in Sprague–Dawley rats, exposure to high doses of CORT, but not low doses, reduced density of immature neurons and cell proliferation and in the dentate gyrus [@DOSEneurobrummelte2010chronic]. The CORT treatment employed here is known to impact other aspects such as growth or mortality in these species (unpublished data), but the effect of the same dose can differ between tissues (@lemaire_prenatal_2000). More striking is the absence of an effect of incubation temperature since most of the studies in reptiles reported better performance when incubated at higher temperatures in different associative learning and motor tasks [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. Furthermore, those studies that reported cold-incubated lizards learn faster [@dayananda_incubation_2017; @abayarathna_effects_2020] used incubation temperatures far beyond the natural range in their hot treatments. We know that our prenatal treatments affect other traits in both species of lizards used in this experiment (unpublished data). However, it is possible that those treatments affect tissues differently. An alternative hypothesis is that the effects of the prenatal treatments are age-dependent [see @frisone2002social]. Here, the lizards' age at the beginning of the experiment ranged from 41 to 148 days old in _L. delicata_ and 48 to 132 in _L. guichenoti_, and when this variable was included in the models, we did no find any significant effect of age (see Supplementary material). However, the treatments may have consequences at earlier or later stages of life that are not being captured in our experiment. Conversely, the effects of the early environment may likely vary depending on the cognitive task, as evidenced by studies showing variations in the impact of prenatal GCs and temperature on different brain regions [@coomber_independent_1997; @sakata_neural_2000; @lemaire_prenatal_2000]. These hypotheses are not mutually exclusive and could be tested in future experiments using multiple doses of prenatal CORT treatment, assessing learning using multiple cognitive tasks and at different ages, and measuring the neurophysiological effects on different brain nuclei.

@@

```{r, modelsage}
#| label: modelsage
# Fitting the models for L. delicata and L. guichenoti including age as a covariate
#
refit2 = FALSE
formula <- FC_reversal ~ age.start + trial_reversal*cort*temp + (1 + trial_reversal|lizard_id)
species <- c("delicata", "guichenoti")
#
models <- list() # Create an empty list to store the models
for(s in species){
  # Subset the data for the species
  data_species <- clean_df %>%
    filter(species == s)
  # Define model file name
  model_file <- paste0("output/models/model_age_", s, ".rds")
# Fit the model only if it has not been fit yet (if refit2 = TRUE)
  if(refit2){
    # Fit the model
    model <- brm(formula,
                data = data_species,
                family = bernoulli(link = "logit"),
                chains = 4, cores = 4, iter = 3000, warmup = 1000, control = list(adapt_delta = 0.99))
    # Write the model to a file
    saveRDS(model, file = model_file)
  } else {
      # Read the model from a file
      model <- readRDS(file = model_file)
  }
  # Save the model to the list
  models[[s]] <- model
}
model_age_deli <- models[["delicata"]]
model_age_guich <- models[["guichenoti"]]
```

Results for _L. delicata_:
` r summary(model_age_deli)`

Results for _L. guichenoti_:
` r summary(model_age_guich)`
